---
title: "Embedding Computations & Code"
subtitle: "Session 3"
engine: markdown
---

# Session Overview

{{< include /program/_session3.md >}}

## Session Objectives

This session explores Quarto's computational capabilities, focusing on seamless multi-language workflows, code cell options, and integration patterns for R and Python data science stacks.

## Learning Objectives

By the end of this session, participants will be able to:

- Execute R, Python, Julia code.
- Configure code cells using "comment symbol" + "pipe" (`|`) syntax.
- Control code visibility and execution.
- Cache computations for improved performance.

# Computing Environments

## Setup R, Python, and/or Julia for Quarto

:::: {.panel-tabset}

### R

```r
# Install renv if not already installed
install.packages("renv")

# Initialise project
renv::init()

# Install packages
renv::install("rmarkdown")

# Save current state
renv::snapshot()
```

### Python (uv)

```bash
# Install uv (if not installed)
pip install uv

# Initialise project
uv init --no-package --vcs none

# Create virtual environment
uv venv

# Activate environment
source .venv/bin/activate

# Add packages
uv add jupyter jupyter papermill
```

### Julia

```julia
# Start Julia in your project directory
using Pkg

# Activate current directory as project
Pkg.activate(".")

# Add packages
Pkg.add("IJulia")

# Instantiate environment
Pkg.instantiate()
```

:::

## Engine Configuration

Configure engines in your document header:

- `knitr`
- `jupyter`
- `julia`
- `markdown`

:::: {layout-ncol="3"}

```yaml
---
engine: knitr
---
```

```yaml
---
engine: jupyter
jupyter: python3
---
```

```yaml
---
engine: julia
---
```

:::

:::: {.absolute top=150 right=150 width="500" height="300"}

:::: {.callout-tip}
Set explicitly the engine in your document header instead of relying on auto-detection.
:::

:::

## What is a Code Cell?

A **code cell** (also called a **code chunk**) is a section of executable code (defined by ```` ```{language}````) embedded within your Quarto document.

Code cells allow you to:

- Execute programming code directly in your document.
- Display both the code and its output.
- Create reproducible analyses and reports.

## Code Cell Structure

````{.qmd code-line-numbers="|1|2-4|6-14"}
```{python} # <1>
#| label: my-plot # <2>
#| echo: true # <2>
#| fig-cap: "Sample visualization" # <2>

import matplotlib.pyplot as plt # <3>
import numpy as np # <3>
# <3>
x = np.linspace(0, 10, 100) # <3>
y = np.sin(x) # <3>
# <3>
plt.plot(x, y) # <3>
plt.title("Sine Wave") # <3>
plt.show() # <3>
```
````

1. Computing language (*e.g.*, `python`, `r`, `julia`).
2. Code cell options (*e.g.*, `label`, `echo`, `eval`).
3. Code content (your actual programming code).

## Why Use Code Cells?

:::: {.columns}

:::: {.column}

### Reproducibility

- Code and results stay together.
- Easy to re-run analyses.
- Version control friendly.

:::

:::: {.column}

### Documentation

- Combine narrative and code.
- Show methodology clearly.
- Create living documents.

:::

:::

# Execution Options

## Setting Execution Options Globally {.smaller}

```yaml
execute:
  ...
```

+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Option       | Description                                                                                                                                                                                       |
+==============+===================================================================================================================================================================================================+
| `eval`       | Evaluate the code chunk (if `false`, just echos the code into the output).                                                                                                                        |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `echo`       | Include the source code in output                                                                                                                                                                 |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `output`     | Include the results of executing the code in the output (`true`, `false`, or `asis` to indicate that the output is raw markdown and should not have any of Quarto's standard enclosing markdown). |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `warning`    | Include warnings in the output.                                                                                                                                                                   |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `error`      | Include errors in the output (note that this implies that errors executing code will not halt processing of the document).                                                                        |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `include`    | Catch all for preventing any output (code or results) from being included (e.g. `include: false` suppresses all output from the code block).                                                      |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `renderings` | Specify rendering names for the plot or table outputs of the cell, e.g. `[light, dark]`                                                                                                           |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

## Code Cells Options

The "comment symbol" + "pipe" (`|`) syntax allows to pass options to code cells.

**Code Visibility Control**:

````{.qmd code-line-numbers="|2|3|4|5"}
```{r}
#| echo: false        # <1>
#| eval: false        # <2>
#| include: false     # <3>
#| code-fold: true    # <4>
```
````

1. Hide code, show output.
2. Show code, don't execute.
3. Execute code, hide everything.
4. Collapsible code blocks.

:::: {.content-visible when-format="revealjs"}

## Code Cells Options

:::

**Output Control**:

````{.qmd code-line-numbers="|2|3|4|5"}
```{r}
#| warning: false     # <1>
#| message: false     # <2>
#| error: true        # <3>
#| output: false      # <4>
```
````

1. Suppress warnings
2. Suppress messages
3. Show errors for debugging
4. Hide all output

# Cache and Freeze

## Two Systems

:::: {.columns}

:::: {.column}

### Cache: Stores computation results

- Speeds up development.
- Works during iteration.
- Engine-specific behaviour.

:::

:::: {.column}

### Freeze: Prevents re-execution

- Enables collaboration.
- Project-wide control.
- Version control friendly.

:::

:::

## Cache: Development Speed

Store expensive computation results to avoid re-running during development.

Two Types:

- **Jupyter Cache:** Document-level (all-or-nothing).
- **Knitr Cache:** Cell-level (granular control).

```yaml
execute:
  cache: true
```

## Freeze: Collaboration Control

Prevent documents from re-executing during project renders.

Results saved in `_freeze` directory (commit to version control).

```yaml
execute:
 freeze: auto    # Re-render only when source changes
 freeze: true    # Never re-render during project render
```

## Key Behavioural Differences

| Action                     | Cache              | Freeze                         |
|----------------------------|--------------------|--------------------------------|
| Individual document render | ✅ Used            | ❌ Ignored                      |
| Project render             | ✅ Used            | ✅ Controls execution           |
| Prose changes              | No effect          | No effect                      |
| Code changes               | Invalidates cache  | Triggers re-render (auto mode) |

## When to Use Each

:::: {.columns}

:::: {.column}

### Use Cache When

- Developing single format documents.
- Iterating on analysis.
- Working with expensive computations.
- Editing prose frequently.

:::

:::: {.column}

### Use Freeze When

- Working in teams.
- Managing large projects.
- Environment setup is complex.
- Results need to be reproducible.

:::

:::

## Essential CLI Commands

### Cache Control

```bash
quarto render --cache           # Force cache use
quarto render --no-cache        # Disable cache
quarto render --cache-refresh   # Force refresh
```

### Freeze Management

```bash
quarto render document.qmd      # Always executes (ignores freeze)
quarto render                   # Respects freeze settings
```

## Golden Rules

1. Cache for Development.  
   Use when iterating on individual documents

2. Freeze for Collaboration.  
   Use when working with teams or managing large projects

3. Different Scopes.
   - Cache: Session-level performance
   - Freeze: Project-level workflow management

4. Version Control.  
   Always commit `_freeze` directory contents

# Parameters

## What Are Parameters?

Parameters allow you to create **dynamic variations** of the same document:

- **Geographic reports** → Different locations
- **Time-based analysis** → Different periods  
- **Scenario modelling** → Different assumptions
- **Multi-client reports** → Different datasets

Think of parameters as **variables** that make your documents reusable and flexible.

## Defining Parameters

:::: {.panel-tabset}

### Python (`papermill`)

```python
#| tags: [parameters]
alpha = 0.1
ratio = 0.5
country = "UK"
year = 2024
```

- Must use `#| tags: [parameters]` comment.
- Provide sensible default values.
- Parameters available in top-level environment.
- Works with `.qmd` and `.ipynb` files.

### R (`knitr`)

```yaml
---
title: "Sales Report"
params:
  region: "North"
  start_date: "2024-01-01"
  threshold: 1000
---
```

Access in R code:

```r
print(params[["region"]])
```

:::

## Passing Parameters: Command Line

Use the `-P` (`--execute-param`) flag to override parameter values:

```bash
# Single parameter
quarto render report.qmd -P country:France

# Multiple parameters  
quarto render report.qmd -P year:2023 -P region:Europe

# With output specification
quarto render report.qmd -P country:Germany --to typst
```

## Passing Parameters: YAML Files

Create reusable parameter sets with YAML files:

:::: {.columns}

:::: {.column}

```{.yaml filename="params.yml"}
country: "Spain"
year: 2023
budget: 50000
```

:::

:::: {.column}

```{.bash filename="Terminal"}
quarto render report.qmd \
  --execute-params params.yml
```

:::

:::

**Benefits:** Version control, team sharing, complex configurations.

---

## Practical Example {.smaller}

:::: {.columns}

:::: {.column}

### Setup (Jupyter/Python)

```python
#| tags: [parameters]
region = "All"
quarter = "Q1" 
min_sales = 1000
```

### Using parameters

```python
if region != "All":
   df = df[df['region'] == region]

df_filtered = df[df['sales'] >= min_sales]
title = f"{region} Sales - {quarter}"
```

:::

:::: {.column}

### Command line usage

```bash
quarto render sales.qmd \
 -P region:North \
 -P quarter:Q2
```

:::

:::

**Result:** Customised report for North region, Q2 data.

## Best Practices

:::: {.columns}

:::: {.column}

### ✅ Do

- Provide meaningful defaults.
- Use descriptive names.
- Document parameter types.
- Validate values in code.
- Use YAML for complex configs.
:::

::: {.column}

### ❌ Avoid

- Generic names (`x`, `val`).
- Parameters without defaults.
- Hardcoded values.
- Overly complex structures.
:::

::::

## Workflow Integration

### Automated reporting

```bash
# Daily reports
quarto render report.qmd -P date:"$(date +%Y-%m-%d)"

# Multiple outputs
for region in North South East West; do
 quarto render regional.qmd -P region:$region
done
```

**Perfect for:** CI/CD pipelines, batch processing, scheduled reports.

{{< include _exercises.qmd >}}

:::: {.content-visible when-format="revealjs"}
{{< include /_extensions/mcanouil/coeos/_about.qmd >}}
:::
