---
title: "Embedding Computations & Code"
subtitle: "Session 3"
engine: markdown
code-links:
  - text: Exercises (tar.gz)
    icon: file-code
    href: assets/exercises/03-exercises.tar.gz
---

# Session Overview

{{< include /program/_session3.md >}}

## Session Objectives

This session explores Quarto's computational capabilities, focusing on seamless multi-language workflows, code cell options, and integration patterns for R and Python data science stacks.

## Learning Objectives

::: {.fragment fragment-index="1"}

By the end of this session, participants will be able to:

:::

- [Execute R {{< iconify fa7-brands:r-project >}}, Python {{< iconify fa7-brands:python >}}, Julia {{< iconify simple-icons:julia >}} code.]{.fragment fragment-index="2"}
- [Configure code cells using "comment symbol" + "pipe" (`|`) syntax.]{.fragment fragment-index="3"}
- [Control code visibility and execution.]{.fragment fragment-index="4"}
- [Cache computations for improved performance.]{.fragment fragment-index="5"}

# Computing Environments

## Setup R, Python, and/or Julia for Quarto

::: {.panel-tabset}

### R

```r
# Install renv if not already installed
install.packages("renv")

# Initialise project
renv::init()

# Install packages
renv::install("rmarkdown")

# Save current state
renv::snapshot()
```

### Python (uv)

```bash
# Install uv (if not installed)
pip install uv

# Initialise project
uv init --no-package --vcs none

# Create virtual environment
uv venv

# Activate environment
source .venv/bin/activate

# Add packages
uv add jupyter jupyter papermill
```

### Julia

```julia
# Start Julia in your project directory
using Pkg

# Activate current directory as project
Pkg.activate(".")

# Add packages
Pkg.add("IJulia")

# Instantiate environment
Pkg.instantiate()
```

:::

## Engine Configuration

::: {.fragment fragment-index="1"}

Configure engines in your document header:

:::

- [`knitr` {{< iconify fa7-brands:r-project >}}]{.fragment fragment-index="2"}
- [`jupyter` {{< iconify fa7-brands:python >}}]{.fragment fragment-index="3"}
- [`julia` {{< iconify simple-icons:julia >}}]{.fragment fragment-index="4"}

::: {layout-ncol="3" style="margin-top: 2em;"}

::: {.fragment fragment-index="2"}

```yaml
---
engine: knitr
---
```

:::

::: {.fragment fragment-index="3"}

```yaml
---
engine: jupyter
jupyter: python3
---
```

:::

::: {.fragment fragment-index="4"}

```yaml
---
engine: julia
---
```

:::

:::

::: {.absolute top=175 right=150 width="500" height="300"}

::: {.callout-tip .fragment fragment-index="5"}

Set explicitly the engine in your document header instead of relying on auto-detection which does not work for inline code (*i.e.*, `` `{language} ...` ``).

:::

:::

## What is a Code Cell?

::: {.fragment fragment-index="1"}

A **code cell** (also called a **code chunk**) is a section of executable code (defined by ```` ```{language}````) embedded within your Quarto document.

:::

::: {.fragment fragment-index="2"}

Code cells allow you to:

:::

- [Execute programming code directly in your document.]{.fragment fragment-index="2"}
- [Display both the code and its output.]{.fragment fragment-index="3"}
- [Create reproducible analyses and reports.]{.fragment fragment-index="4"}

## Code Cell Structure

````{.qmd code-line-numbers="|1|2-4|6-14|"}
```{python} # <1>
#| label: my-plot # <2>
#| echo: true # <2>
#| fig-cap: "Sample visualization" # <2>

import matplotlib.pyplot as plt # <3>
import numpy as np # <3>
# <3>
x = np.linspace(0, 10, 100) # <3>
y = np.sin(x) # <3>
# <3>
plt.plot(x, y) # <3>
plt.title("Sine Wave") # <3>
plt.show() # <3>
```
````

1. Computing language (*e.g.*, `python`, `r`, `julia`).
2. Code cell options (*e.g.*, `label`, `echo`, `eval`).
3. Code content (your actual programming code).

## Why Use Code Cells?

::: {layout-ncol="2"}

::: {.fragment fragment-index="1"}

### Reproducibility

- Code and results stay together.
- Easy to re-run analyses.
- Version control friendly.

:::

::: {.fragment fragment-index="2"}

### Documentation

- Combine narrative and code.
- Show methodology clearly.
- Create living documents.

:::

:::

# Execution Options

## Setting Execution Options Globally {.smaller}

```yaml
execute:
  ...
```

+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Option       | Description                                                                                                                                                                                       |
+==============+===================================================================================================================================================================================================+
| `eval`       | Evaluate the code chunk (if `false`, just echos the code into the output).                                                                                                                        |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `echo`       | Include the source code in output                                                                                                                                                                 |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `output`     | Include the results of executing the code in the output (`true`, `false`, or `asis` to indicate that the output is raw markdown and should not have any of Quarto's standard enclosing markdown). |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `warning`    | Include warnings in the output.                                                                                                                                                                   |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `error`      | Include errors in the output (note that this implies that errors executing code will not halt processing of the document).                                                                        |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `include`    | Catch all for preventing any output (code or results) from being included (e.g., `include: false` suppresses all output from the code block).                                                     |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `renderings` | Specify rendering names for the plot or table outputs of the cell, e.g., `[light, dark]`. See [Renderings](https://quarto.org/docs/computations/renderings.html).                                 |
+--------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: `execute` options {#tbl-execute-options tbl-colwidths="[25,75]" style="font-size: 85%; margin-top: 1em;"}

## Code Cells Options

::: {.fragment fragment-index="1"}

The "comment symbol" + "pipe" (`|`) syntax allows to pass options to code cells.

:::

::: {.fragment fragment-index="2"}

**Code Visibility Control**:

````{.qmd code-line-numbers="|2|3|4|5|"}
```{r}
#| echo: false        # <1>
#| eval: false        # <2>
#| include: false     # <3>
#| code-fold: true    # <4>
```
````

1. Hide code, show output.
2. Show code, don't execute.
3. Execute code, hide everything.
4. Collapsible code blocks.

:::

::: {.content-visible when-format="revealjs"}

## Code Cells Options

:::

**Output Control**:

````{.qmd code-line-numbers="|2|3|4|5|"}
```{r}
#| warning: false     # <1>
#| message: false     # <2>
#| error: true        # <3>
#| output: false      # <4>
```
````

1. Suppress warnings
2. Suppress messages
3. Show errors for debugging
4. Hide all output

# Cache and Freeze

## Two Systems

::: {layout-ncol="2"}

::: {.fragment fragment-index="1"}

### Cache: Stores computation results

- Speeds up development.
- Works during iteration.
- Engine-specific behaviour.

:::

::: {.fragment fragment-index="2"}

### Freeze: Prevents re-execution

- Enables collaboration.
- Project-wide control.
- Version control friendly.

:::

:::

## Cache: Development Speed

::: {.fragment fragment-index="1"}

Store expensive computation results to avoid re-running during development.

:::

::: {.fragment fragment-index="2"}

```yaml
execute:
  cache: true
```

:::

::: {.fragment fragment-index="3"}

Two Types:

:::

- [**Jupyter Cache:** Document-level (*all-or-nothing*).]{.fragment fragment-index="4"}
- [**Knitr Cache:** Cell-level (*granular control*).]{.fragment fragment-index="5"}

## Freeze: Collaboration Control

::: {.fragment fragment-index="1"}

Prevent documents from re-executing during project renders.

:::

::: {.fragment fragment-index="2"}

Results saved in `_freeze` directory (commit to version control).

:::

::: {.fragment fragment-index="3"}

```yaml
execute:
 freeze: auto    # Re-render only when source changes
 freeze: true    # Never re-render during project render
```

:::

## Key Behavioural Differences {auto-animate=true}

| Action                     | Cache              | Freeze                         |
|----------------------------|--------------------|--------------------------------|
| Individual document render | ✅ Used            | ❌ Ignored                      |
| Project render             | ✅ Used            | ✅ Controls execution           |
| Prose changes              | No effect          | No effect                      |
| Code changes               | Invalidates cache  | Triggers re-render (auto mode) |

## When to Use Each

::: {layout-ncol="2"}

::: {.fragment fragment-index="1"}

### Use Cache When

- Developing multi format documents.
- Iterating on analysis.
- Working with expensive computations.
- Editing prose frequently.

:::

::: {.fragment fragment-index="2"}

### Use Freeze When

- Working in teams.
- Managing large projects.
- Environment setup is complex.
- Results need to be reproducible.

:::

:::

## Essential CLI Commands

::: {.fragment fragment-index="1"}

### Cache Control

```bash
quarto render --cache           # Force cache use
quarto render --no-cache        # Disable cache
quarto render --cache-refresh   # Force refresh
```

:::

::: {.fragment fragment-index="2"}

### Freeze Management

```bash
quarto render document.qmd      # Always executes (ignores freeze)
quarto render                   # Respects freeze settings
```

:::

## Golden Rules {.smaller}

::: {.fragment fragment-index="1"}

1. **Cache for Development.**  
   Use when iterating on individual documents.

:::

::: {.fragment fragment-index="2"}

2. **Freeze for Collaboration.**  
   Use when working with teams or managing large projects.

:::

::: {.fragment fragment-index="3"}

3. **Different Scopes.**
   - Cache: Session-level performance.
   - Freeze: Project-level workflow management.

:::

::: {.fragment fragment-index="4"}

4. **Version Control.**  
   Always commit `_freeze` directory contents.

:::

# Parameters

## What Are Parameters?

::: {.fragment fragment-index="1"}

Parameters allow you to create **dynamic variations** of the same document:

:::

- [**Geographic reports** → Different locations]{.fragment fragment-index="2"}
- [**Time-based analysis** → Different periods]{.fragment fragment-index="3"}
- [**Scenario modelling** → Different assumptions]{.fragment fragment-index="4"}
- [**Multi-client reports** → Different datasets]{.fragment fragment-index="5"}

::: {.callout-tip .fragment fragment-index="6"}

Think of parameters as **variables** that make your documents code reusable and flexible.

:::

## Defining Parameters

::: {.panel-tabset}

### Python {{< iconify fa7-brands:python >}} (`papermill`)

```python
#| tags: [parameters]
alpha = 0.1
ratio = 0.5
country = "UK"
year = 2024
```

- Must use `#| tags: [parameters]` comment.
- Provide sensible default values.
- Parameters available in top-level environment.
- Works with `.qmd` and `.ipynb` files.

### R {{< iconify fa7-brands:r-project >}} (`knitr`)

```yaml
---
title: "Sales Report"
params:
  region: "North"
  start_date: "2024-01-01"
  threshold: 1000
---
```

Access in R code:

```r
print(params[["region"]])
```

:::

## Passing Parameters: Command Line

::: {.fragment fragment-index="1"}

Use the `-P` (`--execute-param`) flag to override parameter values:

:::

::: {.fragment fragment-index="2"}

```bash
# Single parameter
quarto render report.qmd --execute-param country:France

# Multiple parameters  
quarto render report.qmd -P year:2023 -P region:Europe

# With output specification
quarto render report.qmd -P country:Germany --to typst
```

:::

## Passing Parameters: YAML Files

::: {.fragment fragment-index="1"}

Create reusable parameter sets with YAML files:

:::

::: {layout-ncol="2"}

::: {.fragment fragment-index="2"}

```{.yaml filename="params.yml"}
country: "Spain"
year: 2023
budget: 50000
```

:::

::: {.fragment fragment-index="3"}

```{.bash filename="Terminal"}
quarto render report.qmd \
  --execute-params params.yml
```

:::

:::

::: {.callout-tip .fragment fragment-index="4"}

## Benefits

Version control, team sharing, complex configurations.

:::

## Practical Example {.smaller}

::: {layout-ncol="2"}

::: {}

::: {.fragment fragment-index="1"}

### Setup (Jupyter/Python)

````qmd
```{{python}}
#| tags: [parameters]
region = "All"
quarter = "Q1" 
min_sales = 1000
```
````

:::

::: {.fragment fragment-index="2"}

### Using parameters

````qmd
```{{python}}
if region != "All":
   df = df[df['region'] == region]

df_filtered = df[df['sales'] >= min_sales]
title = f"{region} Sales - {quarter}"
```
````

:::

:::

::: {}

::: {.fragment fragment-index="3"}

### Command line usage

```bash
quarto render sales.qmd \
  --execute-param region:North \
  --execute-param quarter:Q2
```

:::

::: {.callout-tip .fragment fragment-index="4"}

## Results

Customised report for North region, Q2 data.

:::

:::

:::

## Best Practices

::: {layout-ncol="2"}

::: {}

::: {.fragment fragment-index="1"}

### ✅ Do

:::

- [Provide meaningful defaults.]{.fragment fragment-index="2"}
- [Use descriptive names.]{.fragment fragment-index="3"}
- [Document parameter types.]{.fragment fragment-index="4"}
- [Validate values in code.]{.fragment fragment-index="5"}
- [Use YAML for complex configs.]{.fragment fragment-index="6"}

:::

::: {}

::: {.fragment fragment-index="7"}

### ❌ Avoid

:::

- [Generic names (`x`, `val`).]{.fragment fragment-index="8"}
- [Parameters without defaults.]{.fragment fragment-index="9"}
- [Hardcoded values.]{.fragment fragment-index="10"}
- [Overly complex structures.]{.fragment fragment-index="11"}

:::

:::

## Workflow Integration

::: {.fragment fragment-index="1"}

### Automated reporting

```bash
# Daily reports
quarto render report.qmd -P date:"$(date +%Y-%m-%d)"

# Multiple outputs
for region in North South East West; do
  quarto render regional.qmd -P region:$region
done
```

:::

::: {.callout-tip .fragment fragment-index="2"}

## Perfect for

CI/CD pipelines, batch processing, scheduled reports.

:::

# Hands-On Exercise: Adding Computational Power to Your Portfolio

## Exercise Overview

**Objective**: Transform your Session 2 portfolio document into a computational showcase by adding live code execution, demonstrating the power of embedding computations in Quarto documents using modern Python tools.

**Example Code**: [{{< iconify octicon:file-code-16 >}} Exercises](assets/exercises/03-exercises.tar.gz)

```bash
tar -xzf "03-exercises.tar.gz" -C "03-embedding-computations-code"
```

## Part 1: Environment Setup {.smaller}

Choose your language and update your YAML header:

::: {.panel-tabset}

### Python {{< iconify fa7-brands:python >}}

```yaml
---
title: "My Computational Portfolio"
author: "Your Name"
engine: jupyter
jupyter: python3
execute:
  echo: true
  warning: false
---
```

Install packages:

```bash
uv add polars plotnine # pip install polars plotnine
```

### R {{< iconify fa7-brands:r-project >}}

```yaml
---
title: "My Computational Portfolio"
author: "Your Name"
engine: knitr
execute:
  echo: true
  warning: false
---
```

Install packages:

```r
install.packages(c("dplyr", "ggplot2"))
```

### Julia {{< iconify simple-icons:julia >}}

```yaml
---
title: "My Computational Portfolio"
author: "Your Name"
engine: julia
execute:
  echo: true
  warning: false
---
```

Install packages:

```julia
using Pkg
Pkg.add(["DataFrames", "Plots", "StatsPlots"])
```

:::

## Part 2: Replace Static with Dynamic {.smaller}

**Replace your skills table from Session 2**:

::: {.panel-tabset}

### Python {{< iconify fa7-brands:python >}}

- Create a Polars `DataFrame` with your skills data.
- Include columns: skill name, confidence level, hours practiced.
- Use `print()` to display the data nicely.
- Calculate total practice hours with `pl.sum()`.
- Show the result using inline code `` `{python} your_calculation` ``.

### R {{< iconify fa7-brands:r-project >}}

- Create a `data.frame` with your skills data.
- Include columns: skill name, confidence level, hours practiced.
- Use `knitr::kable()` to display the table nicely.
- Calculate total practice hours with `sum()`.
- Show the result using inline code `` `{r} your_calculation` ``.

### Julia {{< iconify simple-icons:julia >}}

- Create a `DataFrame` with your skills data.
- Include columns: skill name, confidence level, hours practiced.
- Display the `DataFrame` directly.
- Calculate total practice hours with `sum()`.
- Show the result using inline code `` `{julia} your_calculation` ``.

:::

## Part 3: Create Your First Visualization {.smaller}

**Make a chart of your skills**:

::: {.panel-tabset}

### Python {{< iconify fa7-brands:python >}}

- Use plotnine to create a bar chart.
- Try `geom_col()` or `geom_bar()`.
- Add proper labels with `labs()`.
- Use `theme_minimal()` for clean styling.
- Add a figure caption with `#| fig-cap:`.

### R {{< iconify fa7-brands:r-project >}}

- Use ggplot2 to create a bar chart.
- Try `geom_col()` or `geom_bar()`.
- Add proper labels with `labs()`.
- Use `theme_minimal()` for clean styling.
- Add a figure caption with `#| fig-cap:`.

### Julia {{< iconify simple-icons:julia >}}

- Use Plots.jl to create a bar chart.
- Try `bar()` function.
- Add proper labels with `xlabel!()` and `ylabel!()`.
- Add a title with `title!()`.
- Add a figure caption with `#| fig-cap:`.

:::

## Part 4: Experiment with Code Visibility {.smaller}

**Try different code cell options**:

::: {.panel-tabset}

### Python {{< iconify fa7-brands:python >}}

- Create a cell with `#| eval: false` (shows code, doesn't run).
- Make a collapsible section with `#| code-fold: true`.
- Hide code with `#| echo: false` but show output.
- Use `#| include: false` for data setup.

### R {{< iconify fa7-brands:r-project >}}

- Create a cell with `#| eval: false` (shows code, doesn't run).
- Make a collapsible section with `#| code-fold: true`.
- Hide code with `#| echo: false` but show output.
- Use `#| include: false` for library loading.

### Julia {{< iconify simple-icons:julia >}}

- Create a cell with `#| eval: false` (shows code, doesn't run).
- Make a collapsible section with `#| code-fold: true`.
- Hide code with `#| echo: false` but show output.
- Use `#| include: false` for package imports.

:::

## Success Criteria

✅ **You've successfully completed the exercise if you can**:

- Transform static content into dynamic, computed content using Python {{< iconify fa7-brands:python >}}, R {{< iconify fa7-brands:r-project >}}, or Julia {{< iconify simple-icons:julia >}}.
- Create and display data visualisations.
- Use inline code to insert computed values into prose.
- Experiment with different code cell visibility options.
