---
title: "Introduction & Setup"
subtitle: "Session 1"
---

# Introduction

:::: {.content-visible when-format="revealjs"}
## About Me

:::: {style="font-size: 50%;"}
{{< include /assets/quarto/_about-social.qmd >}}
:::
:::: {style="font-size: 55%; font-style: italic;"}
{{< include /assets/quarto/_about-description.qmd >}}
:::
:::

{{< include /program/_session1.md >}}

```{r}
#| label: setup-screenshots-directory
#| include: false
dir.create("assets/screenshots", recursive = TRUE, showWarnings = FALSE)
```

## Workshop Objectives

- **Familiarise** participants with the Quarto ecosystem.
- **Empower** attendees to create reproducible documents and interactive publications.
- **Guide** learners through embedding code, adjusting outputs, and publishing projects.
- **Encourage** hands‑on practice with interactive sessions.

# What is {{< quarto >}} Quarto? {{< iconify fluent-emoji:thinking-face >}}

:::: {.content-visible when-format="revealjs"}

## What is {{< quarto >}} Quarto? {{< iconify fluent-emoji:thinking-face >}}

:::

:::: {#quarto-introduction}
[{{< quarto >}} Quarto](https://quarto.org/) is an open-source scientific and technical publishing system built on Pandoc.  
It is a next-generation R Markdown and Jupyter Notebook interface that allows you to create dynamic documents, reports, presentations, and websites.
:::

:::: {style="display: none;"}
:::: {#quarto-screenshot style="text-align:center;"}

```{r}
#| label: screenshot-quarto-org
webshot2::webshot(
  url = "https://quarto.org/",
  file = "assets/screenshots/quarto-org.png",
  vwidth = 1920 * 0.55,
  vheight = 1080 * 0.55,
  cliprect = "viewport"
)
```

:::

:::: {#quarto-description}

{{< quarto >}} Quarto ([quarto.org](https://quarto.org/)):

- Open-source scientific and technical publishing system built on Pandoc.
- Multi-language platform supporting Python, R, Julia, and Observable.js.
- Creates dynamic, reproducible documents from code and narrative text.
- Single-source authoring for articles, websites, books, presentations, and dashboards.

:::
:::

:::: {.content-visible unless-format="revealjs"}
:::: {layout="[0.30, 0.70]" layout-valign="center"}

:::: {reuse="quarto-screenshot"}
:::

:::: {reuse="quarto-description"}
:::

:::
:::

:::: {.content-visible when-format="revealjs"}
:::: {style="max-width: 50%;"}
:::: {reuse="quarto-screenshot"}
:::
:::

:::: {.attribution}
<https://quarto.org/>
:::

## Purpose

:::: {reuse="quarto-description"}
:::
:::

## Key Strengths

Multi-format output
: Generate HTML, PDF, Word, ePub, and more from one source.

Advanced markdown
: Cross-references, equations, citations, callouts, and complex layouts.

Tool flexibility
: Works with VS Code, RStudio, JupyterLab, or any text editor.

Project system
: Render document collections with shared configurations.

Interactive elements
: Embed widgets, Shiny apps, and Observable JS for dynamic content.

## Reproducibility Benefits

Embedded computation
: Code and results stay together, eliminating copy-paste errors.

Version control friendly
: Plain text format works seamlessly with Git.

Transparent workflows
: Complete audit trail from data to final output.

Collaborative research
: Shared environments and dependencies ensure consistent results.

## What is Pandoc? {{< iconify fluent-emoji:thinking-face >}}

:::: {layout="[0.30, 0.70]" layout-valign="center"}

```{r}
#| label: screenshot-pandoc-org
webshot2::webshot(
  url = "https://pandoc.org/",
  file = "assets/screenshots/pandoc-org.png",
  vwidth = 1920 * 0.60,
  vheight = 1080 * 1.80,
  cliprect = "viewport"
)
```

:::: {#pandoc-description}
Pandoc ([pandoc.org](https://pandoc.org/)):

- is a **free software**, released under the GPL by [John MacFarlane](http://johnmacfarlane.net/).
- is a **command line interface** (CLI).
- allows to **convert to and between multiple formats**.
- **understands Markdown** syntax.
- **understands LaTeX math** and macros.
- handles **citations and bibliographies**.
:::
:::

## How does {{< quarto >}} Quarto work? {{< iconify fluent-emoji:thinking-face >}}

```{mermaid}
%%| label: quarto-workflow
%%| fig-responsive: false
%%| fig-height: 6
%%{
  init: {
    "theme": "dark",
    "themeCSS": ".label foreignObject, .cluster-label foreignObject { font-size: 90%; overflow: visible; }"
  }
}%%
flowchart LR
  A1[qmd] --> B1{"R<br>(<code>engine: knitr</code>)"}
  A1[qmd] --> B2{"Python<br>(<code>engine: jupyter</code>)"}
  A1[qmd] --> B3{"Julia<br>(<code>engine: julia</code>)"}
  A2[ipynb] --> B2
  B1 --> C[md]
  B2 --> C[md]
  B3 --> C[md]
  C --> D{"Pandoc"}
  D --> E["HTML"]
  D --> F["Typst"]
  D --> G["Reveal.js"]
  D --> H[...]

  subgraph engine [Engine]
  B1
  B2
  B3
  end
```

## A Guide to {{< quarto >}} Quarto Versions

{{< quarto >}} Quarto releases uses a versioning scheme similar to [Semantic Versioning](https://semver.org/) with the following release types:

:::: {layout-ncol="3"}
Pre-release

Release Candidate (RC)[^release-candidate]

Release
:::

[^release-candidate]: A release candidate is a pre-release version of the software that is almost ready for release but may still have some minor bugs or issues to be resolved.

```{mermaid}
%%| label: quarto-versions
timeline
    section Release
      1.0 : July, 2022
      1.1 : August, 2022
      1.2 : November, 2022
      1.3 : April, 2023
      1.4 : Fall 2023
      1.5 : Summer 2024
      1.6 : Fall 2024
      1.7 : Spring 2025 Current
    section Pre-Release
      1.8
```

# Getting started

## Downloading and installing Quarto

```{r}
#| label: screenshot-quarto-download
#| output: asis
page <- "https://quarto.org/docs/download/"
file <- "assets/screenshots/quarto-download.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.75,
  vheight = 1080 * 0.75,
  cliprect = "viewport"
)
cat(sprintf("\n\n[![](%s){width=100%% fig-align=center fig-alt='Screenshot of %s webpage'}](%s)", file, page, page))
```

:::: {.attribution}
<https://quarto.org/>
:::

:::: {.content-visible when-format="revealjs"}
## Downloading and installing Quarto
:::

Additionally, you can download and install Quarto using:

:::: {.columns}

:::: {.column}

- [Homebrew](https://brew.sh/) (macOS/linux)

```{.bash filename="Bash"}
brew install quarto
```

```{r}
#| label: screenshot-quarto-homebrew
#| output: asis
page <- "https://brew.sh/"
file <- "assets/screenshots/quarto-homebrew.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n[![<a href = '%s'>%s</a>](%s){fig-alt='Screenshot of %s webpage'}](%s)", page, page, file, page, "https://formulae.brew.sh/cask/quarto"))
cat("\n\n:::\n\n")
```

:::

:::: {.column}

- [Chocolatey](https://chocolatey.org/) (Windows)

```{.PowerShell filename="PowerShell"}
choco install quarto
```

```{r}
#| label: screenshot-quarto-chocolatey
#| output: asis
page <- "https://community.chocolatey.org/"
file <- "assets/screenshots/quarto-chocolatey.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n[![<a href = '%s'>%s</a>](%s){fig-alt='Screenshot of %s webpage'}](%s)", page, page, file, page, "https://community.chocolatey.org/packages/quarto"))
cat("\n\n:::\n\n")
```

:::
:::

## {{< quarto >}} Quarto: a command line interface

```{bash}
#| label: quarto-command-line
#| echo: true
#| attr-output: 'style="font-size: 28%;"'
#| output-location: fragment
quarto --help
```

## Checking {{< quarto >}} Quarto is installed {.smaller}

```bash
quarto check
```

```{.sh code-line-numbers="|1-12|14-16|18-21|23-26|30-34|38-45|"}
Quarto 1.7.32
[✓] Checking environment information...                  # <1>
      Quarto cache location: /home/vscode/.cache/quarto  # <1>
[✓] Checking versions of quarto binary dependencies...   # <1>
      Pandoc version 3.6.3: OK                           # <1>
      Dart Sass version 1.85.1: OK                       # <1>
      Deno version 1.46.3: OK                            # <1>
      Typst version 0.13.0: OK                           # <1>
[✓] Checking versions of quarto dependencies......OK     # <1>
[✓] Checking Quarto installation......OK                 # <1>
      Version: 1.7.32                                    # <1>
      Path: /opt/quarto/bin                              # <1>

[✓] Checking tools....................OK                 # <2>
      TinyTeX: (external install)                        # <2>
      Chromium: (not installed)                          # <2>

[✓] Checking LaTeX....................OK                 # <3>
      Using: TinyTex                                     # <3>
      Path: /home/vscode/.TinyTeX/bin/x86_64-linux       # <3>
      Version: 2025                                      # <3>

[✓] Checking Chrome Headless....................OK       # <4>
      Using: Chrome found on system                      # <4>
      Path: /usr/bin/google-chrome                       # <4>
      Source: PATH                                       # <4>

[✓] Checking basic markdown render....OK

[✓] Checking Python 3 installation....OK                 # <5>
      Version: 3.13.5                                    # <5>
      Path: /usr/local/python/current/bin/python3        # <5>
      Jupyter: 5.8.1                                     # <5>
      Kernels: julia-1.11, python3                       # <5>

[✓] Checking Jupyter engine render....OK

[✓] Checking R installation...........OK                 # <6>
      Version: 4.5.1                                     # <6>
      Path: /opt/R/4.5.1/lib/R                           # <6>
      LibPaths:                                          # <6>
        - /home/vscode/R/x86_64-pc-linux-gnu-library/4.5 # <6>
        - /opt/R/4.5.1/lib/R/library                     # <6>
      knitr: 1.50                                        # <6>
      rmarkdown: 2.29                                    # <6>

[✓] Checking Knitr engine render......OK
```

1. Quarto version and dependencies.
2. Tools installed by Quarto (*e.g.*, `quarto install`).
3. LaTeX installation detected by Quarto.
4. Chromium-based browser installation detected by Quarto.
5. Python installation and computing dependencies detected by Quarto.
6. R installation and computing dependencies detected by Quarto.

# {{< quarto >}} Quarto Projects

## Creating a new {{< quarto >}} Quarto project

```{.bash filename="Bash"}
quarto create project
```

:::: {.fragment}

```{.bash}
 ? Type
 ❯ default
   website
   blog
   manuscript
   book
   confluence
```

:::

## {{< quarto >}} Quarto Projects: `default`

:::: {layout-ncol="2" layout-valign="center"}

:::: {}

```{.bash filename="Bash"}
quarto create project default Default
```

```{bash}
#| label: quarto-project-default-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Default
```

:::

:::: {.fragment}

```{r}
#| label: screenshot-quarto-project-default
#| output: asis
system("cd _demo/Default; quarto render Default.qmd --quiet")
page <- "_demo/Default/Default.html"
file <- "assets/screenshots/quarto-project-default.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

:::: {.content-visible when-format="revealjs"}
## {{< quarto >}} Quarto Projects: `default`
:::

- `_quarto.yml`: project configuration file.

```{r}
#| label: quarto-project-default-yml
#| output: asis
cat(
  '\n\n```{.yaml filename="Quarto"}',
  readLines("_demo/Default/_quarto.yml"),
  "```\n\n",
  sep = "\n"
)
```

- `Default.qmd`: default Quarto document.

```{r}
#| label: quarto-project-default-qmd
#| output: asis
cat(
  '\n\n```{.markdown filename="Quarto"}',
  readLines("_demo/Default/Default.qmd"),
  "```\n\n",
  sep = "\n"
)
```

## {{< quarto >}} Quarto Projects: `website`

:::: {layout-ncol="2" layout-valign="center"}

:::: {}

```{.bash filename="Bash"}
quarto create project website Website
```

```{bash}
#| label: quarto-project-website-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Website
```

:::

:::: {.fragment}

```{r}
#| label: screenshot-quarto-project-website
#| output: asis
system("cd _demo/Website; quarto render index.qmd --quiet")
page <- "_demo/Website/_site/index.html"
file <- "assets/screenshots/quarto-project-website.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n:::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `blog`

:::: {layout-ncol="2" layout-valign="center"}

:::: {}

```{.bash filename="Bash"}
quarto create project blog Blog
```

```{bash}
#| label: quarto-project-blog-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Blog
```

:::

:::: {.fragment}

```{r}
#| label: screenshot-quarto-project-blog
#| output: asis
system("cd _demo/Blog; quarto render index.qmd --quiet")
page <- "_demo/Blog/_site/index.html"
file <- "assets/screenshots/quarto-project-blog.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n:::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `book`

:::: {layout-ncol="2" layout-valign="center"}

:::: {}

```{.bash filename="Bash"}
quarto create project book Book
```

```{bash}
#| label: quarto-project-book-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Book
```

:::

:::: {.fragment}

```{r}
#| label: screenshot-quarto-project-book
#| output: asis
system("cd _demo/Book; quarto render index.qmd --quiet")
page <- "_demo/Book/_book/index.html"
file <- "assets/screenshots/quarto-project-book.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n:::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `manuscript`

:::: {layout-ncol="2" layout-valign="center"}

:::: {}

```{.bash filename="Bash"}
quarto create project manuscript Manuscript
```

```{bash}
#| label: quarto-project-manuscript-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Manuscript
```

:::

:::: {.fragment}

```{r}
#| label: screenshot-quarto-project-manuscript
#| output: asis
system("cd _demo/Manuscript; quarto render index.qmd --quiet")
page <- "_demo/Manuscript/_manuscript/index.html"
file <- "assets/screenshots/quarto-project-manuscript.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n:::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## Writing with your favourite editor {.smaller}

```{r}
#| label: screenshot-quarto-editor
#| classes: panel-tabset
#| output: asis
pages <- sprintf(
  "https://quarto.org/docs/get-started/hello/%s",
  c(
    "vscode.html",
    "jupyter.html",
    "rstudio.html",
    "neovim.html",
    "text-editor.html"
  )
)
for (ipage in pages) {
  file <- paste0(
    "assets/screenshots/quarto-get-started-",
    sub("\\.html$", ".png", basename(ipage))
  )
  webshot2::webshot(
    url = ipage,
    file = file,
    vwidth = 1920 * 1.2,
    vheight = 1080 * 1,
    cliprect = "viewport"
  )
  cat("\n\n###", toupper(sub("\\.html$", "", basename(ipage))))
  cat(sprintf("\n\n[![](%s){fig-alt='Screenshot of %s webpage'}](%s)", file, ipage, ipage))
}
```

:::: {.attribution}
<https://quarto.org/>
:::

## Writing using the Visual Editor {.smaller}

```{r}
#| label: screenshot-quarto-visual-editor
#| output: asis
#| classes: panel-tabset
pages <- c(
  "vscode" = "https://quarto.org/docs/visual-editor/vscode/",
  "rstudio" = "https://quarto.org/docs/visual-editor/"
)
for (ipage in names(pages)) {
  file <- sprintf(
    "assets/screenshots/quarto-visual-editor-%s.png",
    ipage
  )
  webshot2::webshot(
    url = pages[[ipage]],
    file = file,
    vwidth = 1920 * 1.2,
    vheight = 1080 * 1,
    cliprect = "viewport"
  )
  cat("\n\n##", toupper(ipage))
  cat(sprintf("\n\n[![](%s){fig-alt='Screenshot of %s webpage'}](%s)", file, pages[[ipage]], pages[[ipage]]))
}
```

:::: {.attribution}
<https://quarto.org/>
:::

{{< include _exercises.qmd >}}

:::: {.content-visible when-format="revealjs"}
{{< include /_extensions/mcanouil/coeos/_about.qmd >}}
:::
