---
title: "Introduction & Setup"
subtitle: "Session 1"
engine: knitr
code-links:
  - text: Exercises (tar.gz)
    icon: file-code
    href: assets/exercises/01-exercises.tar.gz
---

# Introduction

::: {.content-visible when-format="revealjs"}
## About Me

::: {style="font-size: 50%;"}
{{< include /assets/quarto/_about-social.qmd >}}
:::
::: {style="font-size: 55%; font-style: italic;"}
{{< include /assets/quarto/_about-description.qmd >}}
:::
:::

{{< include /program/_session1.md >}}

```{bash}
#| label: setup-screenshots-directory
#| include: false
rm -rf _demo;
mkdir -p _demo;
mkdir -p assets/screenshots;
```

## Workshop Objectives

::: {.incremental}

- **Familiarise** participants with the Quarto ecosystem.
- **Empower** attendees to create reproducible documents and interactive publications.
- **Guide** learners through embedding code, adjusting outputs, and publishing projects.
- **Encourage** hands-on practice with interactive sessions.

:::

## Learning Objectives

::: {.fragment fragment-index="1"}

By the end of this session, participants will be able to:

:::

- [Explain what Quarto is and how it relates to Pandoc.]{.fragment fragment-index="2"}
- [Install Quarto and verify the installation.]{.fragment fragment-index="3"}
- [Create and render different Quarto project types.]{.fragment fragment-index="4"}
- [Set up an IDE for Quarto development.]{.fragment fragment-index="5"}

# What is {{< quarto >}} Quarto? {{< iconify fluent-emoji:thinking-face >}}

::: {.content-visible when-format="revealjs"}

## What is {{< quarto >}} Quarto? {{< iconify fluent-emoji:thinking-face >}}

:::

::: {style="display: none;"}
::: {#quarto-screenshot style="text-align:center;"}

```{r}
#| label: screenshot-quarto-org
#| output: asis
file <- "assets/screenshots/quarto-org.png"
res <- webshot2::webshot(
  url = "https://quarto.org/",
  file = file,
  vwidth = 1920 * 0.55,
  vheight = 1080 * 0.55,
  cliprect = "viewport"
)
cat(sprintf("\n\n![](%s){fig-align=center width='50%%' fig-alt='Screenshot of Quarto.org home page'}", file))
```

:::

::: {#quarto-description}

[{{< quarto >}} Quarto](https://quarto.org/) is an open-source scientific and technical publishing system built on Pandoc:

- Multi-language platform supporting Python, R, Julia, and Observable.js.
- Creates dynamic, reproducible documents from code and narrative text.
- Single-source authoring for articles, websites, books, presentations, and dashboards.

:::
:::

::: {.content-visible unless-format="revealjs"}
::: {layout="[0.30, 0.70]" layout-valign="center"}

::: {reuse="quarto-screenshot"}
:::

::: {reuse="quarto-description"}
:::

:::
:::

::: {.content-visible when-format="revealjs"}
::: {reuse="quarto-screenshot"}
:::

::: {.attribution}
<https://quarto.org/>
:::

## Purpose

::: {reuse="quarto-description" .incremental}
:::
:::

## Key Strengths {.smaller}

::: {.fragment}

Multi-format output
: Generate HTML, PDF, Word, ePub, and more from one source.

:::
::: {.fragment}

Advanced markdown
: Cross-references, equations, citations, callouts, and complex layouts.

:::
::: {.fragment}

Tool flexibility
: Works with VS Code, RStudio, JupyterLab, or any text editor.

:::
::: {.fragment}

Project system
: Render document collections with shared configurations.

:::
::: {.fragment}

Interactive elements
: Embed widgets, Shiny apps, and Observable JS for dynamic content.

:::

## Reproducibility Benefits {.smaller}

::: {.fragment}

Embedded computation
: Code and results stay together, eliminating copy-paste errors.

:::
::: {.fragment}

Version control friendly
: Plain text format works seamlessly with Git.

:::
::: {.fragment}

Transparent workflows
: Complete audit trail from data to final output.

:::
::: {.fragment}

Collaborative research
: Shared environments and dependencies ensure consistent results.

:::

## What is Pandoc? {{< iconify fluent-emoji:thinking-face >}}

::: {layout="[0.30, 0.70]" layout-valign="center"}

```{r}
#| label: screenshot-pandoc-org
#| output: asis
file <- "assets/screenshots/pandoc-org.png"
res <- webshot2::webshot(
  url = "https://pandoc.org/",
  file = file,
  vwidth = 1920 * 0.60,
  vheight = 1080 * 1.80,
  cliprect = "viewport"
)
cat(sprintf("\n\n![](%s){fig-align=center fig-alt='Screenshot of Pandoc.org home page'}", file))
```

::: {#pandoc-description}
Pandoc ([pandoc.org](https://pandoc.org/)):

- is a **free software**, released under the GPL by [John MacFarlane](http://johnmacfarlane.net/).
- is a **command line interface** (CLI).
- allows to **convert to and between multiple formats**.
- **understands Markdown** syntax.
- **understands LaTeX math** and macros.
- handles **citations and bibliographies**.
:::
:::

## How does {{< quarto >}} Quarto work? {{< iconify fluent-emoji:thinking-face >}}

```{mermaid}
%%| label: quarto-workflow
%%| fig-responsive: false
%%| fig-width: 6
%%| fig-alt: "Quarto workflow diagram"
---
init:
  theme: "dark"
  themeCSS: ".label foreignObject, .cluster-label foreignObject { font-size: 90%; overflow: visible; }"
---
flowchart LR
  A1[qmd] --> B1{"R<br>(<code>engine: knitr</code>)"}
  A1[qmd] --> B2{"Python<br>(<code>engine: jupyter</code>)"}
  A1[qmd] --> B3{"Julia<br>(<code>engine: julia</code>)"}
  A2[ipynb] --> B2
  B1 --> C[md]
  B2 --> C[md]
  B3 --> C[md]
  C --> D{"Pandoc"}
  D --> E["HTML"]
  D --> F["Typst"]
  D --> G["Reveal.js"]
  D --> H[...]

  subgraph engine [Engine]
  B1
  B2
  B3
  end
```

## A Guide to {{< quarto >}} Quarto Versions

{{< quarto >}} Quarto follows a versioning scheme similar to [Semantic Versioning](https://semver.org/) with three release types: Pre-release, Release Candidate (RC), and Release.

::: {.callout-tip}

Always use the latest stable release for production work.
See the [Quarto Changelog](https://quarto.org/docs/download/) for the full release history.

:::

# Getting started

## Downloading and installing Quarto

```{r}
#| label: screenshot-quarto-download
#| output: asis
page <- "https://quarto.org/docs/download/"
file <- "assets/screenshots/quarto-download.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.75,
  vheight = 1080 * 0.75,
  cliprect = "viewport"
)
cat(sprintf("\n\n[![](%s){width=100%% fig-align=center fig-alt='Screenshot of %s webpage'}](%s)", file, page, page))
```

::: {.attribution}
<https://quarto.org/>
:::

::: {.content-visible when-format="revealjs"}
## Downloading and installing Quarto
:::

Additionally, you can download and install Quarto using:

::: {layout-ncol="2"}

::: {}

```{.bash filename="Bash"}
brew install quarto
```

```{r}
#| label: screenshot-quarto-homebrew
#| output: asis
page <- "https://brew.sh/"
file <- "assets/screenshots/quarto-homebrew.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf(
  "\n\n[![<a href = '%s'>%s</a>](%s){fig-alt='Screenshot of %s webpage'}](%s)",
  page,
  page,
  file,
  page,
  "https://formulae.brew.sh/cask/quarto"
))
cat("\n\n:::\n\n")
```

:::

::: {}

```{.PowerShell filename="PowerShell"}
choco install quarto
```

```{r}
#| label: screenshot-quarto-chocolatey
#| output: asis
page <- "https://community.chocolatey.org/"
file <- "assets/screenshots/quarto-chocolatey.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf(
  "\n\n[![<a href = '%s'>%s</a>](%s){fig-alt='Screenshot of %s webpage'}](%s)",
  page,
  page,
  file,
  page,
  "https://community.chocolatey.org/packages/quarto"
))
cat("\n\n:::\n\n")
```

:::
:::

## Using GitHub Codespaces

[![](/assets/images/badge.svg){fig-align="center" fig-alt="Open in GitHub Codespaces" height="2em"}](https://codespaces.new/mcanouil/mastering-quarto-cli?quickstart=1&devcontainer_path=.devcontainer%2Fdevcontainer.json)

A pre-configured development environment is available via GitHub Codespaces with all required tools pre-installed: Quarto, R, Python, Julia, TinyTeX, and supporting utilities.

## {{< quarto >}} Quarto: a command line interface

```{bash}
#| label: quarto-command-line
#| echo: true
#| attr-output: 'style="font-size: 20%;"'
quarto --help
```

## Checking {{< quarto >}} Quarto is installed {.smaller}

```bash
quarto check
```

```{.sh code-line-numbers="|1-12|14-16|18-21|23-26|30-34|38-45"}
Quarto 1.8.24
[✓] Checking environment information...                  # <1>
      Quarto cache location: /home/vscode/.cache/quarto  # <1>
[✓] Checking versions of quarto binary dependencies...   # <1>
      Pandoc version 3.6.3: OK                           # <1>
      Dart Sass version 1.87.0: OK                       # <1>
      Deno version 2.3.1: OK                             # <1>
      Typst version 0.13.0: OK                           # <1>
[✓] Checking versions of quarto dependencies......OK     # <1>
[✓] Checking Quarto installation......OK                 # <1>
      Version: 1.8.24                                    # <1>
      Path: /opt/quarto/bin                              # <1>

[✓] Checking tools....................OK                 # <2>
      TinyTeX: (external install)                        # <2>
      Chromium: (not installed)                          # <2>

[✓] Checking LaTeX....................OK                 # <3>
      Using: TinyTex                                     # <3>
      Path: /home/vscode/.TinyTeX/bin/x86_64-linux       # <3>
      Version: 2025                                      # <3>

[✓] Checking Chrome Headless....................OK       # <4>
      Using: Chrome found on system                      # <4>
      Path: /usr/bin/google-chrome                       # <4>
      Source: PATH                                       # <4>

[✓] Checking basic markdown render....OK

[✓] Checking Python 3 installation....OK                 # <5>
      Version: 3.13.7                                    # <5>
      Path: /usr/local/python/current/bin/python3        # <5>
      Jupyter: 5.8.1                                     # <5>
      Kernels: julia-1.11, python3                       # <5>

[✓] Checking Jupyter engine render....OK

[✓] Checking R installation...........OK                 # <6>
      Version: 4.5.1                                     # <6>
      Path: /opt/R/4.5.1/lib/R                           # <6>
      LibPaths:                                          # <6>
        - /home/vscode/R/x86_64-pc-linux-gnu-library/4.5 # <6>
        - /opt/R/4.5.1/lib/R/library                     # <6>
      knitr: 1.50                                        # <6>
      rmarkdown: 2.29                                    # <6>

[✓] Checking Knitr engine render......OK
```

1. Quarto version and dependencies.
2. Tools installed by Quarto (*e.g.*, `quarto install`).
3. LaTeX installation detected by Quarto.
4. Chromium-based browser installation detected by Quarto.
5. Python installation and computing dependencies detected by Quarto.
6. R installation and computing dependencies detected by Quarto.

# {{< quarto >}} Quarto Projects

## Creating a new {{< quarto >}} Quarto project

```{.bash filename="Bash"}
quarto create project
```

::: {.fragment}

```{.bash}
 ? Type
 ❯ default
   website
   blog
   manuscript
   book
   confluence
```

:::

## {{< quarto >}} Quarto Projects: `default`

```{bash}
#| label: quarto-create-demo-default
#| include: false
(
  cd _demo;
  rm -rf Default;
  quarto create project default Default --no-open --no-prompt;
)
```

::: {layout-ncol="2" layout-valign="center"}

::: {}

```{.bash filename="Bash"}
quarto create project default Default
```

```{bash}
#| label: quarto-project-default-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Default
```

```{bash}
#| label: quarto-render-demo-default
#| include: false
(
  cd _demo/Default;
  quarto render;
)
```

:::

::: {.fragment}

```{r}
#| label: screenshot-quarto-project-default
#| output: asis
page <- "_demo/Default/Default.html"
file <- "assets/screenshots/quarto-project-default.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

::: {.content-visible when-format="revealjs"}
## {{< quarto >}} Quarto Projects: `default`
:::

::: {.fragment}

- `_quarto.yml`: project configuration file.

```{r}
#| label: quarto-project-default-yml
#| output: asis
cat(
  '\n\n```{.yaml filename="Quarto"}',
  readLines("_demo/Default/_quarto.yml"),
  "```\n\n",
  sep = "\n"
)
```

:::

::: {.fragment}

- `Default.qmd`: default Quarto document.

```{r}
#| label: quarto-project-default-qmd
#| output: asis
cat(
  '\n\n```{.markdown filename="Quarto"}',
  readLines("_demo/Default/Default.qmd"),
  "```\n\n",
  sep = "\n"
)
```

:::

## {{< quarto >}} Quarto Projects: `website`

```{bash}
#| label: quarto-create-demo-website
#| include: false
(
  cd _demo;
  rm -rf Website;
  quarto create project website Website --no-open --no-prompt;
)
```

::: {layout-ncol="2" layout-valign="center"}

::: {}

```{.bash filename="Bash"}
quarto create project website Website
```

```{bash}
#| label: quarto-project-website-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Website
```

```{bash}
#| label: quarto-render-demo-website
#| include: false
(
  cd _demo/Website;
  quarto render;
)
```

:::

::: {.fragment}

```{r}
#| label: screenshot-quarto-project-website
#| output: asis
page <- "_demo/Website/_site/index.html"
file <- "assets/screenshots/quarto-project-website.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `blog`

```{bash}
#| label: quarto-create-demo-blog
#| include: false
(
  cd _demo;
  rm -rf Blog;
  quarto create project blog Blog --no-open --no-prompt;
)
```

::: {layout-ncol="2" layout-valign="center"}

::: {}

```{.bash filename="Bash"}
quarto create project blog Blog
```

```{bash}
#| label: quarto-project-blog-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Blog
```

```{bash}
#| label: quarto-render-demo-blog
#| include: false
(
  cd _demo/Blog;
  quarto render;
)
```

:::

::: {.fragment}

```{r}
#| label: screenshot-quarto-project-blog
#| output: asis
page <- "_demo/Blog/_site/index.html"
file <- "assets/screenshots/quarto-project-blog.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `book`

```{bash}
#| label: quarto-create-demo-book
#| include: false
(
  cd _demo;
  rm -rf Book;
  quarto create project book Book --no-open --no-prompt;
)
```

::: {layout-ncol="2" layout-valign="center"}

::: {}

```{.bash filename="Bash"}
quarto create project book Book
```

```{bash}
#| label: quarto-project-book-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Book
```

```{bash}
#| label: quarto-render-demo-book
#| include: false
(
  cd _demo/Book;
  quarto render --to html;
)
```

:::

::: {.fragment}

```{r}
#| label: screenshot-quarto-project-book
#| output: asis
page <- "_demo/Book/_book/index.html"
file <- "assets/screenshots/quarto-project-book.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## {{< quarto >}} Quarto Projects: `manuscript`

```{bash}
#| label: quarto-create-demo-manuscript
#| include: false
(
  cd _demo;
  rm -rf Manuscript;
  quarto create project manuscript Manuscript --no-open --no-prompt;
)
```

::: {layout-ncol="2" layout-valign="center"}

::: {}

```{.bash filename="Bash"}
quarto create project manuscript Manuscript
```

```{bash}
#| label: quarto-project-manuscript-tree
#| attr-output: 'style="font-size: 50%;"'
tree -I "_demo" _demo/Manuscript
```

```{bash}
#| label: quarto-render-demo-manuscript
#| include: false
(
  cd _demo/Manuscript;
  quarto render --to html;
)
```

:::

::: {.fragment}

```{r}
#| label: screenshot-quarto-project-manuscript
#| output: asis
page <- "_demo/Manuscript/_manuscript/index.html"
file <- "assets/screenshots/quarto-project-manuscript.png"
res <- webshot2::webshot(
  url = page,
  file = file,
  vwidth = 1920 * 0.70,
  vheight = 1080 * 0.70,
  cliprect = "viewport"
)

cat('\n\n::: {style="text-align:center;"}\n\n')
cat(sprintf("\n\n![](%s){fig-align=center}", file))
cat("\n\n:::\n\n")
```

:::

:::

## Writing with your favourite editor {.smaller}

```{r}
#| label: screenshot-quarto-editor
#| classes: panel-tabset
#| output: asis
pages <- sprintf(
  "https://quarto.org/docs/get-started/hello/%s",
  c(
    "positron.html",
    "vscode.html",
    "jupyter.html",
    "rstudio.html",
    "neovim.html",
    "text-editor.html"
  )
)
for (ipage in pages) {
  file <- paste0(
    "assets/screenshots/quarto-get-started-",
    sub("\\.html$", ".png", basename(ipage))
  )
  webshot2::webshot(
    url = ipage,
    file = file,
    vwidth = 1920 * 1.2,
    vheight = 1080 * 1,
    cliprect = "viewport"
  )
  cat("\n\n###", toupper(sub("\\.html$", "", basename(ipage))))
  cat(sprintf("\n\n[![](%s){fig-align=center fig-alt='Screenshot of %s webpage'}](%s)", file, ipage, ipage))
}
```

::: {.attribution}
<https://quarto.org/>
:::

## Writing using the Visual Editor {.smaller}

```{r}
#| label: screenshot-quarto-visual-editor
#| output: asis
#| classes: panel-tabset
pages <- c(
  "positron" = "https://quarto.org/docs/tools/positron/visual-editor.html",
  "vscode" = "https://quarto.org/docs/visual-editor/vscode/",
  "rstudio" = "https://quarto.org/docs/visual-editor/"
)
for (ipage in names(pages)) {
  file <- sprintf(
    "assets/screenshots/quarto-visual-editor-%s.png",
    ipage
  )
  webshot2::webshot(
    url = pages[[ipage]],
    file = file,
    vwidth = 1920 * 1.2,
    vheight = 1080 * 1,
    cliprect = "viewport"
  )
  cat("\n\n##", toupper(ipage))
  cat(sprintf(
    "\n\n[![](%s){fig-align=center fig-alt='Screenshot of %s webpage'}](%s)",
    file,
    pages[[ipage]],
    pages[[ipage]]
  ))
}
```

::: {.attribution}
<https://quarto.org/>
:::

# Hands-On Exercise: Creating Your First Quarto Project

## Exercise Overview

**Objective**: Create a personal Quarto document that demonstrates basic features, project setup skills, and IDE integration

**Example Code**: [{{< iconify octicon:file-code-16 >}} Exercises](assets/exercises/01-exercises.tar.gz)

```bash
tar -xzf "01-exercises.tar.gz" -C "01-introduction-setup"
```

::: {.callout-tip}

## IDE Configuration

Before starting, ensure your development environment is ready:

- **VS Code**: Install the Quarto extension (`Ctrl/Cmd + Shift + X`, search "Quarto").
- **Positron**: Built-in Quarto support is included (no additional extensions needed).
- Open the Command Palette (`Ctrl/Cmd + Shift + P`) and type "Quarto" to verify available commands.

:::

## Part 1: Project Creation

1. **Choose Your Adventure**: Create a project using the CLI or your IDE.

2. **Explore the project structure** in the IDE's file explorer panel.

## Part 2: IDE Features Exploration {.smaller}

**Discover your IDE's Quarto capabilities**:

- **File recognition**: Notice `.qmd` files have distinctive icons and syntax highlighting.
- **Document outline**: Open a `.qmd` file and check the Outline/Structure panel (shows headings and sections).
- **Quarto assist panel**: Look for a Quarto-specific panel that provides contextual help.
- **Command palette power**: `Ctrl/Cmd + Shift + P` → type "Quarto" to see all available commands.
- **Integrated features**: Visual editor mode, preview panels, terminal integration.
- **Smart editing**: Try auto-completion in YAML headers, fenced divs, and code blocks.

## Part 3: Document Customisation {.smaller}

**Modify the main document** (`index.qmd`) to include:

1. **Update the YAML header** with your information:

   ```{=html}
   <details><summary>Code</summary>
   ```

   ```markdown
   ---
   title: "My Quarto Journey"
   author: "Your Name"
   date: today
   editor: visual  # Enables rich editing features
   ---
   ```

   ```{=html}
   </details>
   ```

2. **Add content** to the document body.

## Part 4: Rendering {.smaller}

**Try different rendering approaches**:

1. **Command Line** (classic method):

   ```bash
   quarto render
   ```

2. **IDE Integration** (modern workflow):

   - Command Palette (`Ctrl/Cmd + Shift + P`) - search for "Quarto Preview" and/or "Quarto Render" commands.
   - Use keyboard shortcut if you configured one
   - Notice IDE-specific feedback and error handling

## Success Criteria {.smaller}

✅ **You've successfully completed the exercise if you can**:

- Configure your IDE for optimal Quarto development.
- Create a new Quarto project using the CLI or your IDE.
- Navigate project files efficiently in your IDE.
- Utilise IDE features like syntax highlighting and command palette.
- Modify documents with IDE assistance (auto-completion, outline view).
- Render documents using both CLI and IDE methods.
